### 2023-04-03 paper-thesis/chapters/20-adding-booleans.tex
Work a bit on thesis
First part of feedback from Jesper
Third part of feedback from Jesper

\begin{lstlisting}
BoolType : Expr
BoolFalse : Expr
BoolTrue : Expr
BoolIf : Expr * Expr * Expr -> Expr
\end{lstlisting}
Then, the rules for beta reducing the language are in figure \ref{fig:bool-rules-beta}. There is one particularly interesting case, that is how to beta-reduce if statements. Converting this rule to Statix is not entirely trivial, since you need to choose which rule to apply based on what \verb|c| evaluates to. A new rule is needed, which has 3 cases. One for if \verb|c| evaluates to true, one for if \verb|c| evaluates to false, and finally a third case for other cases (such as when \verb|c| is a variable that does not have a substitution). These rules are stated below: (Remember that \verb|rebuild| is the rule introduced in chapter~\ref{chap:baselang}, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s)

@@

Next we will add \verb|Postulate| to the language. A postulate declares that there is a variable with a certain type, without specifying a value. Through the view of the Curry-Howard correspondence, this is equivalent to an axiom. At the current stage, this is useful for testing the language. For example, this is a test with a postulate:
\begin{lstlisting}
postulate T : Type;
if true then Bool else T end
\end{lstlisting}
The following rules are used to implement the feature, which translate cleanly to Statix:

@@

\section{Type Assert}

@@

\begin{enumerate}
	\item Define the parsing rules for the new feature
	\item Create a new file, \verb|tp_[feature].stx| and import this file in the main \verb|type_check.stx| file.
	\item In the new file, define a case for the \verb|betaReduceHead| rule for the constructors that were added. Also define cases for \verb|expectBetaEq| and \verb|betaReduce| if necessary. This is only necessary iff the new constructors are not always eliminated by \verb|betaReduceHead|.
	\item In the new file, define a case for the \verb|typeOfExpr| rule for the constructors that were added.
\end{enumerate}

@@

In the following chapters, we're going to be extending the language further with these more interesting features using the approach defined above:
\begin{itemize}
	\item Inference (chapter~\ref{ch:inference})
	\item Inductive Datatypes (chapter~\ref{ch:datatypes})
	\item Universes (chapter~\ref{ch:universes})
\end{itemize}

>>>>

\begin{lstlisting}
BoolType : Expr
BoolFalse : Expr
BoolTrue : Expr
BoolIf : Expr * Expr * Expr -> Expr
\end{lstlisting}
Then, the rules for beta reducing the language are in figure \ref{fig:bool-rules-beta}. There is one particularly interesting case, that is how to beta-reduce if statements. Converting this rule to Statix is not entirely trivial, since you need to choose which rule to apply based on what \verb|c| evaluates to. A new rule is needed, which has 3 cases. One for if \verb|c| evaluates to true, one for if \verb|c| evaluates to false, and finally a third case for other cases (such as when \verb|c| is a variable that does not have a substitution). These rules are stated below: (Remember that \verb|rebuild| is the rule introduced in chapter~\ref{chap:baselang}, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s).

@@

Next we will add \verb|Postulate| to the language. A postulate declares that there is a variable with a certain type, without specifying a value. Through the view of the Curry-Howard correspondence, this is equivalent to an axiom. At the current stage, this is useful for testing the language. For an example of a test with postulates, we assert that this program evaluates to Bool:
\begin{lstlisting}
postulate T : Type;
if true then Bool else T end
\end{lstlisting}
The following rules are used to implement the feature, which translate cleanly to Statix:

@@

\section{Type Assertions}

@@

\begin{enumerate}
	\item Define the parsing rules for the new feature.
	\item Create a new file, \verb|tp_[feature].stx| and import this file in the main \verb|type_check.stx| file.
	\item In the new file, define a case for the \verb|betaReduceHead| rule for the constructors that were added. Also define cases for \verb|expectBetaEq| and \verb|betaReduce| if necessary. This is only necessary if the new constructors are not always eliminated by \verb|betaReduceHead|.
	\item In the new file, define a case for the \verb|typeOfExpr| rule for the constructors that were added.
\end{enumerate}

@@

In the following chapters, we're going to be extending the language further with these more interesting features using the approach defined above:
\begin{itemize}
	\item Inference (chapter~\ref{ch:inference}).
	\item Inductive Datatypes (chapter~\ref{ch:datatypes}).
	\item Universes (chapter~\ref{ch:universes}).
\end{itemize}

### 2023-04-03 paper-thesis/chapters/21-adding-inference.tex
More writing work from today
More writing of the future work
Third part of feedback from Jesper

Inference is an important feature of dependent programming languages, that allows redundant parts of programs to be left out. For example, it allows you to infer the argument type of a function, if it can be inferred by the usage of the function, like in this example where the type of the argument of the polymorphic identity function can be inferred, since we pass it \verb|true| which is a boolean:
\begin{lstlisting}
let id = (\T : Type. \x: T. x);
id _ true
\end{lstlisting}

@@

There are a lot of different algorithms for inference\cite{typeinference}, some algorithms can solve more inferences than others. One algorithm for unification is \emph{first-order unification} (FOU), where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free variable, we set the the value free variable to be equal to the value of the non-free variable. There are some situations in which this approach fails to infer a term, but in most real-world scenarios it works very well. For example, it can infer both programs in the introduction of this chapter, but it fails to infer the following program:

\begin{lstlisting}
let f = _;
\x : Type.
\g: (_: (f x) -> Bool).
g true
\end{lstlisting}

@@

	\item A complex expression \verb|e1| on the left, an infer expression on the right. In this case, we know what top-level constructor of the metavariable should be, but not necessarily the entire constructor (there might be infers in \verb|e1|). We introduce new \verb|Infer| expressions on the left, and call \verb|expectBetaEq| recursively. A simple example for \verb|BoolIf| is below:
	\begin{lstlisting}
expectBetaEq((s1, e1@BoolIf(c1, t1, b1)), (_, Infer(e2))) :-
	e2 == BoolIf(Infer(c2), Infer(t2), Infer(b2)),
	expectBetaEq((s1, e1), (sEmpty(), e2)).
	\end{lstlisting}
	\verb|FnType| and \verb|FnConstruct| work similarly, but we don't only have sub-expressions but also a name to consider. Sadly, this is the first case where we have to approximate. 
	\begin{lstlisting}
expectBetaEq((s1, e1@FnConstruct(arg_name1, arg_type1, body1)), 
	(_, Infer(e2))) :-
	e2 == FnConstruct(arg_name1, Infer(arg_type2), Infer(body2)),
	expectBetaEq_((s1, e1), (sEmpty(), e2)).
	\end{lstlisting}
	\label{inf-case2}
	Ideally, we would generate a new name iff \verb|e2| is a free variable, otherwise using the already generated name. We can't query whether \verb|e2| is free, so as a best-attempt we always assume that \verb|e2| is free and generate a new name, this will fail in some cases. We will discuss this in the next section.

@@

\begin{enumerate}
	\item We could always beta-reduce terms completely to their normal form, then terms that are beta-equal are also identical up to alpha equality. This would solve the problem in case \ref{inf-case4}, but the problem of case \ref{inf-case2} would remain. Furthermore, doing all this beta reduction is terrible for the performance of the type checker: Beta-reducing terms may take an arbitrary amount of time, though it is guaranteed to terminate since the Calculus of Constructions is strongly normalizing~\cite{Coquand_Huet_1988}.
	\item Ideally, we want a declarative definition of the problem. To do this, we need to be able to declare that two terms that are not identical are equal, which is possible under \emph{equational unification}~\cite[Section 2.1]{Siekmann86}. If this system was added to Statix, this would make implementing more powerful inference algorithms cleanly possible.
\end{enumerate}

>>>>

Inference is an important feature of dependent programming languages, that allows redundant parts of programs to be left out. This reduces the annotation burden in comparison with explicitly typed languages. For example, it allows you to infer the argument type of a function, if it can be inferred by the usage of the function, like in this example where the type of the argument of the polymorphic identity function can be inferred, since we pass it \verb|true| which is a boolean:
\begin{lstlisting}
let id = (\T : Type. \x: T. x);
id _ true
\end{lstlisting}

@@

There are a lot of different algorithms for inference\cite{typeinference}, some algorithms can solve more inference problems than others. One algorithm for unification is \emph{first-order unification} (FOU), where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free metavariable, we set the the value free metavariable to be equal to the value of the non-free metavariable. There are some situations in which this approach fails to infer a term, but in most real-world scenarios it works very well. For example, it can infer both programs in the introduction of this chapter, but it fails to infer the following program:

\begin{lstlisting}
let f = _;
\x : Type.
\g: (f x) -> Bool.
g true
\end{lstlisting}

@@

	\item A complex expression \verb|e1| on the left, an infer expression on the right. In this case, we know what top-level constructor of the metavariable should be, but not necessarily the entire constructor (there might be infers in \verb|e1|). We introduce new \verb|Infer| expressions on the left, and call \verb|expectBetaEq| recursively. A simple example for \verb|BoolIf| is below:
	\begin{lstlisting}
expectBetaEq((s1, e1@BoolIf(c1, t1, b1)), (_, Infer(e2))) :-
	e2 == BoolIf(Infer(c2), Infer(t2), Infer(b2)),
	expectBetaEq((s1, e1), (sEmpty(), e2)).
	\end{lstlisting}
	\verb|FnType| and \verb|FnConstruct| work similarly, but we don't only have sub-expressions but also a name to consider. Sadly, this is the first case where we have to approximate. 
	\begin{lstlisting}
expectBetaEq((s1, e1@FnConstruct(arg_name1, arg_type1, body1)), 
	(_, Infer(e2))) :-
	e2 == FnConstruct(arg_name1, Infer(arg_type2), Infer(body2)),
	expectBetaEq_((s1, e1), (sEmpty(), e2)).
	\end{lstlisting}
	\label{inf-case2}
	Ideally, we would generate a new name iff \verb|e2| is a unresolved metavariable, otherwise using the already generated name. We can't query whether \verb|e2| is free, so as a best-attempt we always assume that \verb|e2| is free and generate a new name, this will fail in some cases. We will discuss this in the next section.

@@

\begin{enumerate}
	\item We could always beta-reduce terms completely to their normal form, then terms that are beta-equal are also identical up to alpha equality. This would solve the problem in case \ref{inf-case4}, but the problem of case \ref{inf-case2} would remain. Furthermore, doing all this beta reduction is terrible for the performance of the type checker: Beta-reducing terms may take an arbitrary amount of time, though it is guaranteed to terminate since the Calculus of Constructions is strongly normalizing~\cite{Coquand_Huet_1988}.
	\item Ideally, we want a declarative definition of the problem. To do this, we need to be able to declare that two terms that are not identical are equal, which is possible under \emph{equational unification}~\cite[Section 2.1]{Siekmann86}. If this system was added to Statix, this would make implementing more powerful inference algorithms cleanly possible. This idea is explored further in chapter \ref{ch:comp-lambdapi}.
\end{enumerate}

### 2023-04-03 paper-thesis/chapters/22-adding-datatypes.tex
Update 23 and 24
Third part of feedback from Jesper

A difference between Agda and our implementation is that Agda has case matching as a native construct, whereas we chose to use eliminators as described in Inductive families~\cite{eliminators}. An eliminator is a function that can be used to case match on a data type. For example, the eliminator of the \verb|Nat| type above is:
\begin{lstlisting}
elim Nat : P: (v: Nat -> Type) -> P Z 
	-> (x: Nat -> P x -> P (S x)) -> n: Nat -> P n
\end{lstlisting}

The general type of an eliminator for a data type \verb|N| is:
\begin{lstlisting}
parameters 
-> P: (indices -> v: N params indices -> Type) 
-> constructors
-> indices
-> v: N params indices
-> P indices v
\end{lstlisting}
The meaning of all the arguments is:
\begin{itemize}
	\item \verb|parameters| are the parameters of the data type we want to eliminate, since these are constant among all constructors those should be specified at the start.
	\item \verb|P| is the type that this eliminator will return. It may depend on the value \verb|v| that is eliminated.
	\item \verb|constructors| is a function that eliminates each of the constructors of the datatype. To eliminate a constructor of the form \verb|C : args -> N params indices| it generates a function \verb|args -> P indices (C pars args)|. For recursive datatypes, a previous case is generated, such as the \verb|P x| in the eliminator of \verb|Nat| above.
	\item \verb|indices| are the indices of the value that is to be eliminated.
	\item \verb|v| is the value that is to be eliminated.
	\item Finally, \verb|P indices v| is the result of the elimination.
\end{itemize}

@@

	\item Create a new scope \verb|s4| with \verb|s1| as the parent. Then type-check each constructor with \verb|s2| as scope (so that the constructors can depend on parameters, but cannot depend on indices), and declare the type of that constructor in \verb|s4|. The parameters are automatically added to the type of the constructor.

@@

When defining a recursive data type, we would like our datatype to be \emph{strictly positive}, meaning that it can only be recursive on positive positions. If this rule would not exist, we would be able to create non-terminating functions using the datatype. Since under the curry-howard correspondence this allows to prove false, this is undesirable.

>>>>

A difference between Agda and our implementation is that Agda has pattern matching as a native construct, whereas we chose to use eliminators as described in Inductive families~\cite{eliminators}. An eliminator is a function that can be used to case match on a data type. For example, the eliminator of the \verb|Nat| type above is:
\begin{lstlisting}
elim Nat : P: (v: Nat -> Type) -> P Z 
	-> (x: Nat -> P x -> P (S x)) -> n: Nat -> P n
\end{lstlisting}

The general type of an eliminator for a data type \verb|N| is:
\begin{lstlisting}
parameters 
-> P: (indices -> v: N params indices -> Type) 
-> methods
-> indices
-> v: N params indices
-> P indices v
\end{lstlisting}
The meaning of all the arguments is:
\begin{itemize}
	\item \verb|parameters| are the parameters of the data type we want to eliminate, since these are constant among all constructors those should be specified at the start.
	\item \verb|P| is the type that this eliminator will return. It may depend on the value \verb|v| that is eliminated.
	\item \verb|methods| is for each of the constructors of the datatype, a function that eliminates it. To eliminate a constructor of the form \verb|C : args -> N params indices| it requires a function \verb|args -> P indices (C pars args)|. For recursive datatypes, a previous case is generated, such as the \verb|P x| in the eliminator of \verb|Nat| above.
	\item \verb|indices| are the indices of the value that is to be eliminated.
	\item \verb|v| is the value that is to be eliminated.
	\item Finally, \verb|P indices v| is the result of the elimination.
\end{itemize}

@@

	\item Create a new scope \verb|s4| with \verb|s1| as the parent. Then type-check each constructor with \verb|s2| as scope (so that the constructors can depend on parameters, but cannot depend on indices), and declare the constructor in \verb|s4|. We must also ensure that each constructor returns the datatype it belongs to, with matching parameters, though the indices may of course vary.

@@

When defining a recursive data type, we would like our datatype to be \emph{strictly positive}, meaning that it can only be recursive on positive positions. If this check would not exist, we would be able to create non-terminating functions using the datatype. Since under the Curry Howard correspondence this allows to prove false, this is undesirable.

### 2023-04-03 paper-thesis/chapters/31-comparing-with-lambdapi.tex
Third part of feedback from Jesper

\begin{figure}[ht]
	\begin{lstlisting}
symbol Let : 
	`Π` (A : TmSort), 
	`Π` (B : TmType A → TmSort), 
	`Π` (v : TmType A),
	`Π` (b : TmType A → TmType (B v)),
	TmType (B v);
rule Let _ _ $v $b `$\hookrightarrow$` $b $v;
	\end{lstlisting}
	\caption{Definition of less powerful Let in lampdapi}
	\label{fig:lp-let}
\end{figure}

@@

>>>>

\begin{figure}[ht]
	\begin{lstlisting}
symbol Let : 
	`Π` (A : TmSort), 
	`Π` (B : TmType A → TmSort), 
	`Π` (v : TmType A),
	`Π` (b : TmType A → TmType (B v)),
	TmType (B v);
rule Let _ _ $v $b `$\hookrightarrow$` $b $v;
	\end{lstlisting}
	\caption{Definition of less powerful Let in LambdaPi}
	\label{fig:lp-let}
\end{figure}

@@

\section{Comparison}

In this chapter we've defined the calculus of constructions by embedding it into LambdaPi. This is a declarative definition, similarly to Statix and unlike Haskell. However, since we are embedding the language into LambdaPi, the dynamic semantics of the language are now defined in terms of the dynamic semantics of LambdaPi. This means that the features our language can have are limited by the features of LambdaPi, this means it was impossible to define \verb|Let| bindings the way we wanted to.

The static semantics of the language are defined by the way we create our terms. This is quite flexible, though less flexible than Statix since for example we have no control over the scoping rules (such as shadowing) of the language, which would be controlled by scope graphs in Statix.

The conclusion is that embedding a language in LambdaPi is primarily useful if we want to verify the implementation of our language is correct, which is one of the primary use cases of LambdaPi. Using it as a compiler for the language is not the intention behind LambdaPi, which is clear from the limitations we've mentioned above. We believe this comparison was still insightful, as it shows how compact an implementation in Statix could be if Statix would have equational unification.
